---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
title: Cleaning up the feed from ProMED
author:
- name: Sangeeta Bhatia
  affiliation: Imperial College London
abstract: 
keywords: 
date: "`r Sys.Date()`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# spacing: double
bibliography: 
biblio-style: apsr
endnote: no
---



```{r setup}
library(dplyr)
library(EpiEstim)
library(ggplot2)
library(ggthemes)
library(glue)
library(here)
library(incidence)
library(janitor)
library(lubridate)
library(patchwork)
library(purrr)
library(readr)
library(scales)
library(tidyr)
source(here("R/merge_dup_lines_DS1.R"))
source(here("R/incidence_data_cleanup.R"))
source(here("R/incidence_data_processing.R"))
```


```{r process-ebola-data-1 }
species <- "Humans"
disease <- "Ebola"
wafrica <- c("Sierra Leone", "Liberia", "Guinea")
datasource <- "promed"
outfile <- glue(
  "{Sys.Date()}_{datasource}_loglinear"
)
```

Visualising the raw data.



```{r process-ebola-data-2 }

feed <- read.csv(
  here("data/raw/promed_2014-2016-renamed.csv"),
  stringsAsFactors = FALSE
) %>%
  filter(Species == species &
         Disease == disease &
         Country %in% wafrica)

feed <- clean_names(feed)
## Janitor renames healthmap to health_map. Fix.
feed <- rename(feed,
  healthmap_alert_id = health_map_alert_id
)
```

Strip the date of the time stamp

```{r process-ebola-data-3 }
feed <- separate(
  feed, issue_date,
  sep = " ",
  into = c("issue_date", "time"),
  remove = TRUE
)
```




### Raw data by case category

```{r process-ebola-data-4 }
## Deaths are not used anywhere. No use plotting them
x <- select(feed, issue_date, sc, cc, country) %>%
  gather(
    case_type, count, -c(issue_date, country)
  )

x$issue_date <- as.Date(x$issue_date, format = "%m/%d/%y")
raw_p <- split(x, x$country) %>%
  map(function(x) {
   p <- ggplot(x, aes(issue_date, count, col = case_type)) +
    geom_point(size = 1.1) +
    scale_x_date(date_labels = "%d-%m-%Y") +
    scale_color_manual(
      values = c(sc = "#00BFC4", cc = "#F8766D"),
      breaks = c("sc", "cc"),
      labels = c("Suspected Cases", "Confirmed Cases")
    ) + ylab("Cases") 
  p <- p + theme_minimal() +
    theme(
      panel.border = element_blank(),
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      legend.title = element_blank(),
      legend.position = "top"
    )
   p})

```


# Data clean-up

Split the data by country.
```{r process-ebola-data-5 }
by_location <- split(feed, feed$country)
```

Extract the total case count as a sum of suspected and
confirmed cases.


```{r process-ebola-data-6 }
cum_cases <- map(
  by_location, function(case_count) {
  case_count$date <- mdy(case_count$issue_date)
  case_count$cases <- apply(
    case_count[, c("sc", "cc")], 1, sum_only_na_stays_na
  )
  case_count
})

```

## Merge duplicate alerts

```{r process-ebola-data-7 }
cols_to_keep <- c(
  "location", "country", "disease", "species",
  "healthmap_alert_id", "headline", "url",
  "alert_tag", "feed_name", "lon", "lat"
)
## These are columns we generate ourselves later on
cols_to_keep <- c(cols_to_keep, "date", "cases")

```

```{r process-ebola-data-8 }

no_dups <- map(
  cum_cases, ~ merge_duplicates(., cols_to_keep)
)

```

## Remove outliers. 

```{r process-ebola-data-9 }
use_last <- 20
p.within.k <- 0.50
k_sd <- interval_width_for_p(
  use_last,
  1 - p.within.k
) %>%
  sqrt() %>%
  `[`(2)

no_outliers <- map(
  no_dups, function(df) {
  df <- arrange(df, date)
  df <- filter(df, !is.na(cases)) %>%
    select(date, cases)
  if (nrow(df) == 0) {
    return(NULL)
  } else {
    df <- remove_last_outliers(df,
      use_last = use_last,
      k_sd = k_sd
    )
  }
  message(df$country[1])
  df
  })


```

## Make monotonically increasing

```{r process-ebola-data-10 }

mono_inc <- map(
  no_outliers, make_monotonically_increasing
)

```

## Interpolate missing data
Finally, interpolate missing data.

```{r process-ebola-data-11 }
interpolated <- map(
  mono_inc,
  ~ interpolate_missing_data(., method = "loglinear")
)
```


Check that each dataframe within interpolated corresponds to daily 
series.

```{r process-ebola-data-12 }
map(interpolated, function(x) diff(x$date))
```

## Derive incidence time series from interpolated cumulative cases

Stan is not going to like real numbers for a Poisson process.
So first we cast the cumulative case counts as integers and then 
take the difference.


```{r process-ebola-data-13 }

incid_tall <- map_dfr(interpolated, function(x) {
  x <- select(x,
    date = interpolated_date,
    cases = interpolated_cases
  )
  x$cases <- as.integer(x$cases)
  x$incid <- c(0, diff(x$cases))
  x <- select(x, date, incid)
  x
}, .id = "country")

```

## Plot


### Total cases as sum of sc and cc.

Common plotting logic. x must have a date and a cases column.

```{r process-ebola-data-14 }

cases_ts <- function(x) {
  p <- ggplot(x, aes(date, cases)) +
    geom_point(size = 0.5)
  p <- p + theme_minimal()
  p <- p + xlab("") + ylab("Cases")
  p <- p + scale_x_date(date_labels = "%d-%b-%Y")
  p
}

```

```{r process-ebola-data-15 }
total_p <- map(cum_cases, cases_ts)      
```

### Remove duplicate alerts

```{r process-ebola-data-16 }
nodups_p <- map(no_dups, cases_ts)
nooutss_p <- map(no_outliers, cases_ts)
```
### Make increasing

```{r process-ebola-data-17 }

monoinc_p <- map(mono_inc, cases_ts)

```

### Interpolated data

This needs a slightly different viz.

```{r process-ebola-data-18, eval = TRUE}

interp_p <- map(interpolated, function(x) {
  x$interpolated <- FALSE
  idx <- which(is.na(x$cases))
  x$interpolated[idx] <- TRUE
  x <- select(x,
    date = interpolated_date,
    cases = interpolated_cases,
    interpolated
  )
  x$alpha <- 0.8
  x$alpha[idx] <- 0.4
  p <- ggplot(x, aes(date, cases, alpha = alpha)) + geom_point(size = 0.5, position = position_jitter(width=0.1))
  p <- p + ylab("Cases")
  p <- p + scale_x_date(date_labels = "%d-%b-%Y")
  p <- p +
    scale_alpha_identity()
  p <- p + theme_minimal()
  p <- p + theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(hjust = 0.5),
    panel.border = element_blank(),
    axis.line = element_line(),
    legend.position = "top",
    legend.title = element_blank()    
  )
  p
})

```
## Daily Incidence Series

```{r process-ebola-data-19 }
daily_bycountry <- split(incid_tall, incid_tall$country)

daily_p <- map(
    daily_bycountry,
    function(df) {
      p <- ggplot(df, aes(date, incid)) +
        geom_point(size = 0.5)
      p <- p + ylab("Daily incidence")
      p <- p + theme_minimal()
      p <- p +
        theme(
          axis.title.x = element_blank(),          
          axis.text.x = element_text(hjust = 0.5),
          panel.border = element_blank(),
          axis.line = element_line()
        )
      p <- p + scale_x_date(date_labels = "%d-%b-%Y")
      p

    }    
)

```
Putting them together.

```{r process-ebola-data-20, eval = TRUE}

raw_p[[1]] + total_p[[1]] +
  nodups_p[[1]] + nooutss_p[[1]] +
  monoinc_p[[1]] + interp_p[[1]] +
  daily_p[[1]] +
  plot_layout(ncol = 2)

```

## Processing data collated by the World Health Organization

We will also use the West African Ebola incidence data collated by the WHO
during the epidemic which was made available ~1 year after the end of
the epidemic [here](https://doi.org/10.1098%2Frstb.2016.0308).

The cleaned version of the WHO data consisted of cases reported
between December 2013 and October 2015 in the three most affected
countries - Guinea, Sierra Leone and Liberia. 

The pre-processing steps consist of turning the line count into a
daily incidence series and turning the given tall data into wide data.

The current data set has three possible values for the epidemiological
case definition - confirmed, probable and suspected.
Epidemiological case definition of probable and suspected cases
differed across countries while confirmed cases were the ones
confirmed through lab report. For the purpose of the current analysis,
one approach could be to lump all of them together. The inferred date of
onset (rather than the date reported) is used for estimation.
The columns we use are : Country, EpiCaseDef (probably),
DateOnsetInferred and CL_DistrictRes. 


```{r process-ebola-data-21, eval = TRUE}
who_raw <- read.csv(
  here("data/raw/rstb20160308supp1.csv"),
  colClasses = c(Country = "factor",
                 EpiCaseDef = "character",
                 DateOnsetInferred = "Date",
                 CL_DistrictRes = "factor")) %>%
  select(Country, EpiCaseDef,
         DateOnsetInferred,CL_DistrictRes)

```

## Line-list to incidence time series

This can be easily accomplished using the `incidence` package.

```{r process-ebola-data-22 }
who_incid <- split(who_raw, who_raw$Country) %>%
    map_dfr(~ incidence(.x$DateOnsetInferred) %>%
                as.data.frame,
            .id = "country")
who_incid <- rename(
  who_incid, "date" = "dates",
  incid = "counts"
)
```


## Compare daily incidence from ProMED and WHO

```{r process-ebola-data-23}
daily_incid <- bind_rows(
  WHO = who_incid, ProMED = incid_tall,
  .id = "datasource"
)

ggplot(
  daily_incid, aes(date, incid, col = datasource)
) +
  geom_point(size = 0.5) +
  facet_wrap(~country, ncol = 1, scales = "free_y") +
  scale_x_date(date_labels = "%d-%B-%Y") +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    legend.position = "top",
    legend.title = element_blank()
  )

```

## Estimating reproduction number from ProMED and WHO Data
We will use [EpiEstim](https://doi.org/10.1093/aje/kwt133) to estimate
the effective reproduction number using data from both sources for
each country.
# Parameters for Ebola

Culled from literature.

```{r process-ebola-data-24, eval = TRUE}
mean_SI <- 14.2
CV_SI <- 9.6 / 14.2
SItrunc <- 40
SI_Distr <- sapply(
  0:SItrunc,
  function(e) DiscrSI(e, mean_SI, mean_SI * CV_SI)
)
SI_Distr <- SI_Distr / sum(SI_Distr)
```

```{r process-ebola-data-25}
all_groups <- split(
  daily_incid,
  list(
    daily_incid$datasource,
    daily_incid$country
  )
)

time_window <- 14

all_R <- map_dfr(
  all_groups,
  function(x) {
    I <- x$incid
    start <- 2:(nrow(x) - time_window)
    end <- start + time_window
    res <- EstimateR(
      I,
      T.Start = start,
      T.End = end,
      method = "NonParametricSI",
      SI.Distr = SI_Distr,
      plot = FALSE,
      CV.Posterior = 1,
      Mean.Prior = 1,
      Std.Prior = 0.5
    )
    res$R$date <- x$date[end]
    res$R},
  .id = "source.country"
)

all_R <- separate(
  all_R,
  source.country,
  into = c("Source", "country"),
  extra = "merge",
  fill = "left"
)

```

```{r process-ebola-data-26}
ggplot(all_R) +
  geom_line(aes(date, `Median(R)`, col = Source)) +
  geom_ribbon(
    aes(date, ymin = `Quantile.0.025(R)`,
        ymax = `Quantile.0.975(R)`, fill = Source),
    alpha = 0.4
  ) +
    facet_wrap(~country, ncol = 1, scales = "free_y") +
  scale_x_date(date_labels = "%d-%B-%Y") +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    legend.position = "top",
    legend.title = element_blank()
  )

  
```
